---
title: "Breakout 4: Summarizing Data"
subtitle: "PHUSE US Connect 2021: Introduction to R for the Statistical Programmer"
author: "Mike Stackhouse"
date: "6/9/2021"
output:
  rmarkdown::html_document:
    theme: lumen
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(haven)
library(knitr)
```

## Intro

To get things started with this breakout, let's pick up where we left off from breakout 3!

```{r data_prep} 
# Use haven to read the raw data straight off of GitHub!
raw_dm <- read_xpt(url("https://github.com/phuse-org/TestDataFactory/raw/main/Updated/TDF_SDTM/dm.xpt"))

# Process the raw DM data
dm <- raw_dm %>%
  mutate(
    # Create USUBJID by concatenating STUDYID, SITEID, and SUBJID
    USUBJID = paste0(STUDYID, "-", SITEID, "-", SUBJID),
    # Safety Flag is "Y" if ARM is not Screen Failure
    SAFFL = if_else(ARM != "Screen Failure",
                    "Y",
                    "N"),
    # if/else if/else flow in R
    AGEGR1 = case_when(
      AGE < 65 ~ "<65",
      AGE <= 80 ~ "65-80",
      AGE > 80 ~ ">80",
      TRUE ~ NA_character_
    )
  )

dm %>% 
  select(USUBJID, STUDYID, SITEID, SUBJID, ARM, SAFFL, AGE, AGEGR1) %>% 
  # These calls to `head()` and `kable()` are just to display the data nicely
  head() %>% 
  kable()
```

Now that our data is prepared, we can summarize our data in a few ways. With the `group_by()` function we can specify how we'd like to group our summaries (like any `by` statement in SAS), and the `summarize()` function will control which summaries are performed.

In this breakout, we're going to do two different types of summaries - descriptive statistics and counting. We're going to:

1)  **Summarize** (mean, sd, min, max) the age (AGE) of subjects, across treatment groups (ARM)

2)  **Count** participants age groups (AGEGR1) across treatment groups (ARM)

## Age Summary

Let's start with the Age summary. Age is a numeric variable, and we're going to perform descriptive statistics. For this assignment, do the following:

-   Use the `dm` dataset created above as a starting point
-   **Group** the data by ARM
-   **Summarize** the data by creating 4 new variables: `mean`, `sd`, `min,` and `max`

*Remember - every function that you'll need for this workshop can be found in the [function reference](https://atorus-connect.eastus2.cloudapp.azure.com/phuse_intro_to_r_2021/resources.html) section of the workshop website.*

```{r sum_age}

```

## Age Group Counts

Great! We did some descriptive statistics. But what about dealing with categorical data? For this, instead we'll do some counting.

For this assignment:

-   Once again, **group** the data by treatment group (ARM)
-   **Count** the number of subjects within each age group \*AGEGR1)

*Note: There are multiple ways to solve this one! You'll find with R that quite often, there's multiple ways to solve the same problem. If your organization doesn't have a preferred convention, do what feels most intuitive. And if something feels like it could be more straightforward - it's worth looking into because someone may already made a better approach.*

```{r sum_agegr}

```
