---
title: "Intro to R Demo"
subtitle: "PHUSE US Connect 2021: Introduction to R for the Statistical Programmer"
author: "Mike Stackhouse"
date: "6/9/2021"
output:
  rmarkdown::html_document:
    theme: lumen
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(haven)
library(knitr)
library(Tplyr)
library(pharmaRTF)
```

## Prep Some Data! 

First, let's import some data and derive some variables. We're going to play with some data from the [PHUSE Test Data Factory project's](https://github.com/phuse-org/TestDataFactory) replication of the [CDISC Pilot Data](https://bitbucket.cdisc.org/projects/CED/repos/sdtm-adam-pilot-project/browse). This is an excellent, public source of data following the SDTM v3.2. 

We're going to use the R package [{haven}](https://haven.tidyverse.org/) to read in SAS version 5 Transport files. Better yet - we're going to read it straight off of GitHub! We'll also use [{dplyr}](https://dplyr.tidyverse.org/) to pre-process our data a bit.

```{r data_prep}
# Use haven to read the raw data straight off of GitHub!
raw_dm <- read_xpt(url("https://github.com/phuse-org/TestDataFactory/raw/main/Updated/TDF_SDTM/dm.xpt"))

dm <- raw_dm %>%
  mutate(
    # Create USUBJID by concatenating STUDYID, SITEID, and SUBJID
    USUBJID = paste0(STUDYID, "-", SITEID, "-", SUBJID),
    # Safety Flag is "Y" if ARM is not Screen Failure
    SAFFL = if_else(ARM != "Screen Failure",
                    "Y",
                    "N"),
    # if/else if/else flow in R
    AGEGR1 = case_when(
      AGE < 65 ~ "<65",
      AGE <= 80 ~ "65-80",
      AGE > 80 ~ ">80",
      TRUE ~ NA_character_
    )
  )
```

## Summarize it! 

Next, let's use the Atorus package [{Tplyr}](https://atorus-research.github.io/Tplyr/) to summarize our data and get is ready for display! 

```{r Tplyr}
t <- tplyr_table(dm, ARM, where = SAFFL == "Y") %>% 
  add_layer(
    group_desc(AGE, by = "Age (Years)")
  ) %>% 
  add_layer(
    group_count(AGEGR1, by = "Age Group")
  ) 

dat <- t %>% 
  build() %>% 
  apply_row_masks(row_breaks = TRUE) %>% 
  select(starts_with('row'), var1_Placebo, `var1_Xanomeline Low Dose`, `var1_Xanomeline High Dose`) %>% 
  add_column_headers(
    paste0(" | | Placebo\\line(N=**Placebo**)| Xanomeline Low Dose\\line(N=**Xanomeline Low Dose**)",
             "|Xanomeline High Dose\\line(N=**Xanomeline High Dose**)"),
    header_n = header_n(t)
  )
  
kable(dat)
```

## Style it! 

Starting to look like a table? Yup. Let's use the package [{huxtable}](https://hughjonesd.github.io/huxtable/) to style our table further, much like we'd do with PROC REPORT in SAS&reg;. 

```{r style}
ht <- huxtable::as_hux(dat, add_colnames=FALSE) %>% # `add_colnames` is FALSE because we already added our own
   huxtable::set_bold(1, 1:ncol(dat), TRUE) %>% # Bold the header row
   huxtable::set_align(1, 3:ncol(dat), 'center') %>% # Center align the results headers
   huxtable::set_align(2:nrow(dat), 3:ncol(dat), 'center') %>% # Center align the results
   huxtable::set_valign(1, 1:ncol(dat), 'bottom') %>% # Bottom align the header row
   huxtable::set_bottom_border(1, 1:ncol(dat), 1) %>% # Put a border under the header row
   huxtable::set_top_border(1, 1:ncol(dat), 1) %>% # Put a border above the header row
   huxtable::set_width(1.5) %>% # Set the table width
   huxtable::set_escape_contents(FALSE) %>% # Don't escape RTF syntax
   huxtable::set_col_width(c(.19, .21, .2, .2, .2))

ht
```

## RTF it! 

Finally, let's write the table out into RTF. To finish everything off, we're also goign to need some titles and footnotes. Here we can use another Atorus package named [{pharmaRTF}](https://atorus-research.github.io/pharmaRTF/). 

```{r rtf}
doc <- rtf_doc(ht, header_rows=1) %>%
   # This gets our rows nice and close together - find more info in the docs
   set_ignore_cell_padding(TRUE) %>% 
   set_font_size(9) %>% 
   # Add in a title
   add_titles(
      hf_line("Demographic Summary", bold=TRUE, align = 'left', font_size = 11),
      hf_line("Safety Population", bold=TRUE, align = 'left', font_size = 11)
   ) %>% 
   # Add some footnotes
   add_footnotes(
      hf_line(paste0("Abbreviations:  N = number of subjects in the population; Q1=25th Percentile;", 
                     " Q3= 75th Percentile; SD=Standard Deviation."), bold = TRUE, align='left'),
      hf_line("Number of subjects with non-missing data, used as the denominator.", bold = TRUE, align='left')
   )

# Now we can write out the RTF file! 
write_rtf(doc, file.path("demog1.rtf"))
```

## Let's try something else

Ok - very good. Let's step things up a bit. How about an AE table? Let's add some extra flair as well. All we're going to do here is use a few more of the features available within {Tplyr}. 

```{r ae}
adsl <- read_xpt(url("https://github.com/phuse-org/TestDataFactory/raw/main/Updated/TDF_ADaM/adsl.xpt"))
adae <- read_xpt(url("https://github.com/phuse-org/TestDataFactory/raw/main/Updated/TDF_ADaM/adae.xpt"))

t <- tplyr_table(adae, TRTA, where=SAFFL == "Y" & TRTEMFL == "Y") %>% 
  set_pop_data(adsl) %>% 
  set_pop_treat_var(TRT01A) %>% 
  set_pop_where(SAFFL == "Y") %>% 
  add_total_group() %>% 
  set_count_layer_formats(
    'n_counts' = f_str('xx (xx.x%) xxx', distinct_n, distinct_pct, n)
  ) %>% 
  set_distinct_by(USUBJID) %>% 
  # Add a count layer for overall subjects
  add_layer(
    group_count("Number of patients reporting at least one treatment-emergent adverse event")
   ) %>%
   # Add a nested count layer for AEBODSYS as the outer layer and AEDECOD as the inner layer
   add_layer(
      group_count(vars(AEBODSYS, AEDECOD)) %>% 
         set_distinct_by(USUBJID) %>% 
         # Create nested formatting
         set_nest_count(TRUE) %>% 
         # Alphabetical body system, count for terms
         set_order_count_method(c("byfactor", "bycount")) %>%
         # Create ordering variable from High Dose
         set_ordering_cols("Xanomeline High Dose") %>%
         # Use the number of subjects who had an AE as the order variable
         set_result_order_var(distinct_n)
   ) 

dat <- t %>% 
  build() %>% 
  # Sort the data by layer, body system, then descending occurrence of AEs within body system
  arrange(ord_layer_index, ord_layer_1, desc(ord_layer_2)) %>%
  # Blank out repeating row labels and insert blank rows between layers and body system groups
  apply_row_masks(row_breaks=TRUE, ord_layer_index, ord_layer_1) %>%
  # Pick out columns for display
  select(row_label1, `var1_Placebo`, `var1_Xanomeline Low Dose`,`var1_Xanomeline High Dose`, var1_Total) %>% 
  # Add column headers to the data frame in preparation for styling with huxtable
  add_column_headers(
    paste0("System Organ Class \\line   Preferred Term",
           "| Placebo \\line(N=**Placebo**) \\line n (%) [E]",
           "| Xanomeline \\line Low Dose \\line(N=**Xanomeline Low Dose**) \\line n (%) [E]",
           "| Xanomeline \\line High Dose \\line(N=**Xanomeline High Dose**) \\line n (%) [E]",
           "| Total \\line(N=**Total**) \\line n (%) [E]"
    ), header_n(t)
  )

# Style the table!
ht <- huxtable::as_hux(dat, add_colnames=FALSE) %>% # `add_colnames` is FALSE because we already added our own
   huxtable::set_bold(1, 1:ncol(dat), TRUE) %>% # Bold the header row
   huxtable::set_align(1,2:ncol(dat), 'center') %>% # Center align the results header
   huxtable::set_align(2:nrow(dat), 2:ncol(dat), 'center') %>% # Center align the results
   huxtable::set_valign(1, 1:ncol(dat), 'bottom') %>% # Bottom align the header row
   huxtable::set_bottom_border(1, 1:ncol(dat), 1) %>% # Put a border under the header row
   huxtable::set_top_border(1, 1:ncol(dat), 1) %>% # Put a border above the header row
   huxtable::set_width(1.5) %>% # Set the table width
   huxtable::set_escape_contents(FALSE) %>% # Don't escape RTF syntax
   huxtable::set_col_width(c(.4, .15, .15, .15, .15)) # Set the column widths

# Build the RTF document
doc <- rtf_doc(ht, header_rows=1) %>%
   # This gets our rows nice and close together - find more info in the docs
   set_ignore_cell_padding(TRUE) %>% 
   set_font_size(9) %>% 
   # Add in a title
   add_titles(
      hf_line(paste0("Summary of Treatment-Emergent Adverse Events by Preferred Term in Descending Frequency", 
                      " of Xanomeline High Dose within System Organ Class"), 
                     bold=TRUE, align = 'left', font_size = 11),
      hf_line("Safety Population", bold=TRUE, align = 'left', font_size = 11)
   ) %>%
   # Add some footnotes
   add_footnotes(
      hf_line(paste0("Abbreviations: N = number of subjects in the population; n=number of patients with at least one row event; ",
                     "E=number of events. Percentages are calculated relative to the treatment group N."), 
              bold = TRUE, align='left', font_size = 9),
      hf_line(paste0("Subjects may be counted in more than one row."), 
              bold = TRUE, align='left', font_size = 9)
   )

# Now we can write out the RTF file! 
write_rtf(doc, file.path("ae_teae.rtf"))
```